Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-04-01 index.qmd
Update 4/1

The major assumption for CPH is the proportional hazards (PH) assumption
which states that the ratio of hazard rates for any two subjects must be
constant at all times. This means the effect of the covariates must be
constant over time [@Bustan2018]. To test this assumption, the
hypothesis test below is used: $$
\begin{align*}
H_0&: \text{the predictor's coefficient does vary with time} \\
H_1&: \text{the predictor's coefficient does not vary with time}
\end{align*}
$$

To generate the p-value for the hypothesis test, the scaled Schoenfeld
partial residuals are calculated for each covariate. The Schoenfeld
partial residuals are the difference between the value of the covariate
and the expected value of the covariate at the time of failure
[@klein2005]. To visually confirm the PH assumption, the plot of the
scaled Schoenfeld partial residuals for each covariate should be a
horizontal line [@harrison2021]. The global p–value will provide the
p-value for all covariates in the model which can be compared to an
$\alpha = 0.05$ to determine if the PH assumption has been met for all
covariates [@nahhas2025].

@@

colnames(summarymetrics) <- c("Metric", "Value")
summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
kable(summarymetrics, format = "markdown", padding = 0)

@@

```{r, echo = FALSE}
kable(tidy(cox1), format = "markdown")
g <- glance(cox1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")
```

@@

```{r, echo = FALSE}
plot_model(cox1,
           dot.size = 2,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

@@

@@

@@

The `cox.zph` function can be used to check the proportional hazards
assumption. As mentioned above, the function calculates p values for
each covariate. If the p values are \< .05, the null hypothesis is
rejected indicating the proportional hazards assumption is true. If the
proportional hazards assumption is not true for any covariate, the model
will be recreated without those covariates.

```{r}
cox.zph(cox1)
```

The output above shows the p values for the hypothesis test.

In addition to the hypothesis test, the Shoenfeld residuals calculated
by the `cox.zph()` function are plotted against time below.

@@

The proportional hazards test shows several covariates do not meet the
proportional hazards assumption.

The model is recreated below with another proportional hazards test. To
improve the model and ensure the proportional hazards assumption is met,
several covariates that do not meet the assumption are removed.

@@

cox.zph(cox1.1)

@@

The new model has seven covariates with only one not meeting the
proportional hazards assumption. Although sm5 does not have a p value \>
.05, the assumption overall holds for the new model and sm5 was not
removed from the model because of its large impact on the hazard ratio
shown in the forest plot. The new model's full summary is below.

```{r}
summary(cox1.1)

@@

#### Checking Linearity Assumption

Residuals can first be found by using the `resid(type = "martingale")`
function and passing martingale in the type argument creating a vector
of residuals that can be plotted against the range of values for each
covariate. Example for one predictor

@@

The plots of the martingale residuals vs the range of values for each
covariate is shown above. For each covariate, there is no clear slope
indicating the linearity assumption is true for the covariates in the
model.

#### Checking Independent Censoring

In the dataset being used for this model, the training data does not
contain censored data. The documentation available at the dataset's
source clarifies all engines in the training data have experienced the
desired event of failure at the end of each time series.

survival curve

The survival curve will be plotted by using the survival package's
`survfit()` function. The object created will be plotted using `plot()`

```{r}
survfit <- survfit(cox1.1)

```{r}
ggsurvplot(survfit, data = test, fun = "pct", surv.median.line = "none")
```

```{r}
plot(survfit, mark.time=FALSE, xscale=365.24,
	xlab='Years', ylab='Survival')
lines(survfit[1], lwd=2, xscale=365.24)
```

@@

Hazard Rate:
$h(t|Z) = h_0(t) * e^{B_1Z_1 + B_2Z_2 + B_3Z_3+B_4Z_4+B_5Z_5}$

>>>>

The major assumption for CPH is the proportional hazards (PH) assumption
which states that the ratio of hazard rates for any two subjects must be
constant at all times. This means the effect of the covariates must be
constant over time [@Bustan2018]. To test this assumption, the
hypothesis test below is used: $$
\begin{align*}
H_0&: \text{the log hazard ratio for each covariate remains constant over time.} \\
H_1&: \text{the log hazard ratio for each covariate does not remain constant over time}
\end{align*}
$$

To generate the p-value for the hypothesis test, the scaled Schoenfeld
partial residuals are calculated for each covariate. The Schoenfeld
partial residuals are the difference between the value of the covariate
and the expected value of the covariate at the time of failure
[@klein2005]. If the global p-value is $< 0.05$, than the model has at
least one covariate that violates the PH assumption and each individual
covariate p-value will need to be checked. If the global p-value is
$> 0.05$, then the PH assumption holds for all covariates [@nahhas2025].
To visually confirm the PH assumption, the plot of the scaled Schoenfeld
partial residuals for each covariate should be a horizontal line
[@harrison2021].

@@

colnames(summarymetrics) <- c("Metric", "Value")
summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
kable(summarymetrics, format = "markdown")

@@

```{r}
kable(tidy(cox1), format = "markdown")
g <- glance(cox1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")
```

@@

```{r}
plot_model(cox1,
           dot.size = 2,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@

The survival times for each individual engine in the training data are
independent of each other. The failure of one engine does not impact the
failure of another so this assumption is met.

@@

The training dataset does not contain censored data as the documentation
available at the dataset’s source clarifies all engines in the training
data have experienced the desired event of failure at the end of each
time series. As such, this assumption is met.

@@

To check the linearity assumption, the residuals of the model are needed
which can be generated using the `resid` function in R and assigning the
type to be martingale (`resid(type = "martingale")` ). This will create
a vector of residuals that can be plotted against the range of values
for each covariate.

The plots of the Martingale residuals vs the range of values for each
covariate is shown below. For each covariate, there is no clear slop
indicating the linearity assumption is true for the covariates in the
model.

```{r}
par(mfrow = c(3,3),
      mar = c(2,2,4,1))

martingale <- resid(cox1, type = "martingale")

plot(train$os1, martingale, main = "Martingale Residuals: os1")
abline(h = 0, col = "red")

plot(train$os2, martingale, main = "Martingale Residuals: os2")
abline(h = 0, col = "red")

plot(train$sm1, martingale, main = "Martingale Residuals: sm1")
abline(h = 0, col = "red")

plot(train$sm6, martingale, main = "Martingale Residuals: sm6")
abline(h = 0, col = "red")

plot(train$sm13, martingale, main = "Martingale Residuals: sm13")
abline(h = 0, col = "red")

plot(train$sm14, martingale, main = "Martingale Residuals: sm14")
abline(h = 0, col = "red")

plot(train$sm21, martingale, main = "Martingale Residuals: sm21")
abline(h = 0, col = "red")
```

@@

The `cox.zph` function can be used to check the proportional hazards
assumption. As mentioned above, the function calculates p values for
each covariate. The global p-value will first be compared to
$\alpha = 0.05$. If that p-value is $<\alpha$, then at least one
covariate violates the PH assumption and the model will need to be
adapted.

The output below shows the p-values for the hypothesis test and
Schoenfeld residuals. The global p-value is $< \alpha$ so at least one
covariate violates the PH assumption. Examining each individual
covariate through both p-values and the plot of the Schoenfeld residuals
shows that the covariates os1, os3, sm1, sm5, sm13, and sm14 all violate
the PH assumption.

```{r}
cph_ph <- cox.zph(cox1) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()
kable(cph_ph, format = "markdown")
```

@@

The results of the above assumptions demonstrated that several
covariates violate the PH assumption. As such, the model will be
recreated by removing the covariates that violate the PH assumption.
After recreating the model, the PH assumption and the linearity
assumption will be tested again.

The table below demonstrates the new model information which has seven
covariates: sm2, sm5, sm7, sm11, sm12, sm20, and sm21.

@@

kable(tidy(cox1.1), format = "markdown")
g <- glance(cox1)
g <- g %>% select(n, nevent, concordance, std.error.concordance, logLik, statistic.log, p.value.log, statistic.wald, p.value.wald, statistic.sc, p.value.sc)
kable(g, format = "markdown")

@@

The output below shows the p-values for the new model's PH assumption
hypothesis test and Schoenfeld residuals. The global p-value is
$> \alpha$ so the PH assumption has been met for the model even though
the covariate sm5 has an individual p-value of $< \alpha$. As such, this
new model does not violate the PH assumption and the analysis can
continue.

```{r}
m_2 <- cox.zph(cox1.1) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()

kable(m_2, format = "markdown")

# Function to put ph_plots on one visual
create_ph_plots <- function(coxmodel, row, col){
  # Function takes argument that is a cox model from coxph()

create_ph_plots(cox1.1, 3, 5)

@@

Since a new model was generated, the below plots confirm that the
linearity assumption is still met for this model.

@@

@@
